---
title: "Term Project Proposal"
author: "Vida S. Osei"
date: "5/3/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Problem Statement
Fish is the primary source of animal protein and livelihood in Ghana, particularly for communities living along the coast and communities bordering lagoons and estuaries. Inland fisheries is very important to the fisheries sector in Ghana contributing about 20% of the total fish production to the country. Species exploited in the inland fisheries are mostly mud fish and the tilapias of which Sarotherodon melanotheron, the blackchin Tilapia constitute the majority. Blackchin tilapia plays an important role in the provision of food security and employment to the people of Ghana especially in areas where they exploited mostly.  Yet, very little is known about the status of the stocks and biology of Sarotherodon melanotheron in the Dominli lagoon in the Western region of Ghana. For management strategies employed to manage these stocks to be effective, information on the biology and stock status must be incorporated in the management strategies that will be formulated to manage the stocks. The research questions I will answer are (a) is Sarotherodon melanotheron growing Isometrically in Dominli Lagoon? and (b) is Sarotherodon melanotheron in the Dominli lagoon in good condition?
```{r}

```

#Justification and Significance of Study
Black-chinned tilapia, Sarotherodon melanotheron, belongs to the family Cichlidae in the order Perciformes. It is native to West Africa from Senegal to Zaire and in its native range, it occurs in brackish coastal lagoons and estuaries (Trewavas 1983). This species is plantivorous feeder, consuming mainly algae, detritus and cyanobacteria but also consume small portions of zooplankton (Kone and Teugels 2003, Ayoadze and Ikuala 2007). Cast nets and funnel-shaped entrance fish traps made of wire are the main gears used in exploiting these fish species. They have been introduced to other continents like Asia and America. The cichlids (tilapia) are of high commercial and economic importance to African countries (Abban etal, 2004).  It is one of the major commercially important fish species exploited in brackish water systems in Ghana and constitutes about 59% of fish caught in brackish water systems in Ghana (Welcome, 1972, Blay and Asabere, 1993). The blackchin tilapia fishery provides livelihood to fishers and also serves as cheap source of protein to neighboring communities (Arizi etal, 2014). The biology of this species has been widely studied by different authors. Faunce (2000) in his study on the reproductive biology of Sarotherodon melanotheron in impounded mangrove systems in Florida reported the sex ratio of 1:1 for the species. However, very little is known about the stock status of blackchin tilapia in estuaries and lagoons in Ghana. Few studies have researched Sarotherodon melanotheron in Ghana. Blay (1998) studied the growth and mortality of blackchin tilapia in the Benya lagoon and the Kakum river estuary in Ghana and reported the length at first sexual maturity to be 5.5cm and 4.6 respectively. Other studies have also look at the genetics (Abban etal, 2000) and the ecology of the species (Ofori Danson and Kumi, 2006). Heavy pressure on this fish have led to a reduction in the average size and abundance of this fish in lagoons and estuaries in Ghana. Good management of Sarotherodon melanotheron necessitates adequate information on its abundance, biology and general population structure.  According to King (2007), length-weight relationship and condition factor are important tools for providing basic scientific information to inform management decisions for fish species. Condition factor is specially used in assessing the relative well-being of a fish population (Bolger and Connoly, 1989). Hence, I will estimate the length-weight relationship and the condition factor of Sarotherodon melanotheron in Ghana to understand the well-being of the species and to provide basic scientific data for future stock assessment. 

#Methodology
##Study Site
The area of concern is the Dominli Lagoon located between latitudes 5° 1´, 5o 2´ N and longitudes 2° 44´, 2o 47´ W at Bonyere in the Jomoro district in Ghana. This lagoon covers a total land area of 465,724.4m2. Diverse fish species, crabs, reptiles, molluscs and mangrove is found in this area.

##Fish data
Length and weight data on fish specimens collected monthly from October to September 2012 from the Dominli lagoon will be used for this analysis.  About 50 samples of fish were bought from commercial fishermen and identified to the species level using fish identification manual by Dankwa etal (1999) and Paugy etal (2003 a, b). The specimens were put on ice and then transported to the laboratory for analysis. At the laboratory, the length of the species was taken to the nearest 0.1cm with a measuring board and the weight to the nearest 0.01g using an electronic weight balance. The sex of each fish was also determined.

##Data Analysis
Size distribution, mean weight of fish by month and relationship between length and weight of the fish will be generated for the fish. 

## Length Weight Relatioship
The length-weight model was transformed to a linear model by taking the natural logarithms of both body weight and standard length of the fish. This is important for making the errors additive and stabilizing the variances about the model.
A linear regression analysis was used to fit the relationship between length and body eight of fish. This will be examined using equation.

#####W = aSLb ………. (Blay, 1998)
Where, W = weight of fish specimen (in grammes), SL = standard length of fish (in cm), a = growth constant; and b = exponent of the equation. All plots will be done with the ‘ggplot2’ 
package in the R environment and all analysis will also be done in R.

##Expected Results
I expect that Sarotherodon melanotheron fish species in the Dominli lagoon will have an isometric growth and will be in good condition.



#Reference
Reference
Abban EK, Agyakwah S, Falk TM (2004). Socio-economic importance of Tilapiine fishes. In Abban EK, Casal CMV, Falk TM, Pullin RSV (Eds.), Biodiversity, Management and Utilization of West African Fishes World Fish Center Conference Proceedings. pp. 1-3. 

Ayoade, A.A., and A.O.O. Ikulala. 2007. Length weight relationship, condition factor and stomach contents of Hemichromis bimaculatus, Sarotherodonmelanotheron and 

Chromidotilapia guentheri (Perciformes: Cichlidae) in Eleiyele Lake, southwestern Nigera. Revista de Biología Tropical 55(3-4):969-977.

Blay, J. 1998. Growth and mortality parameters of Sarotherodon melanotheron melanotheron (Teleostei: Cichlidae) in two brackish water systems in Ghana. Ghana Journal of 
Science, 38: 47-55. 

Blay, J & Asabere-Ameyaw, A. 1993. Assessment of the fishery of a stunted population of the cichlid, Sarotherodon melanotheron (Rüppel), in a closed lagoon in Ghana. Journal of
Applied Ichthyology, 9: 1-11.
 
Dankwa, H. R., Abban E. K & Teugels G. G. 1999. Fresh water fishes of Ghana: Identification,
distribution, ecological and economic importance. Annales Sciences of Zoologiques,  283:53. 

Eyeson KN (1979). Studies on egg production, spawning and fry development in Tilapia melanotheron. Ghana J. Sci. 17(1):25-34. 

Eyeson KN (1983). Stunting and reproduction in pond-reared Sarotherodon melanotheron. 
Aquaculture 31:257-267. 

Eyeson KN (1992). Residual bi-parenting oral-brooding in the blackchin tilapia, Sarotherodon melanotheron (Rüppell), J. Fish Biol. 41:145- 146. 

Faunce, C.H. and Lorenz, J.J. 2000. Reproductive biology of the introduced Mayan cichlid, Cichlasoma urophthalmus, within an estuarine mangrove habitat of southern Florida. Environmental Biology Fishes 58: 215-225. 

Kone, T., and G.G. Teugels. 2003. Food habits of brackish water tilapia Sarotherodon melanotheron in riverine and lacustrine environments of a West African coastal basin. 
Hydrobiologia 490:75-85.


Ofori-Danson PK, Kumi GN (2006). Food and feeding habits of Sarotherodon melanotheron 

Rüppell, 1852, (Pisces: Cichlidae) in Sakumo lagoon, Ghana. West Afr. J. Appl. Ecol. 10:21-34.

Paugy, D., Leveque C & Teugels G. G. 2003a. The fresh and brackish water fishes of West Africa, Volume I. Publications Scientifique du Muséum, MRAC, 459p. 


